\documentclass[11pt]{article}

\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{microtype}

\title{The Inspiration Simplex\\[0.4em]
\large Using derivative-free optimization in concept space\\
to create new derivative-free optimizers}
\author{Peter Cotton}
\date{Working paper, \today}

\begin{document}
\maketitle

\begin{abstract}
\noindent Place $K$ established algorithms at the vertices of a simplex. A
point is a vector of mixing weights, and a language model turns the point
into a working program: the heaviest vertex supplies the architecture and
the others graft their ideas into it, in the stated proportions
(Figure~\ref{fig:prompt} shows the literal prompt). Scoring the program on
a benchmark makes the simplex a continuous space of algorithm designs that
ordinary derivative-free optimizers can search. The construction's first
product is Alloy, generated at the equal-weights point: on twenty-nine
problems never used in any selection step it has the best mean rank at
every budget from 60 to 480 evaluations and beats six competitors including
CMA-ES pairwise ($p < 10^{-10}$); it ships in the humpday package. Letting
weights go negative extends the recipes beyond the simplex; of six
semantics for a minus sign, tested under a paired-ablation protocol, one
survives out of sample, with a rule attached: negative weights repair weak
hosts and spoil strong ones. Failures are reported throughout: the
continuous search never beat the obvious point, single generations vary
severalfold, and in-sample scores mislead reliably. Further worked examples
are planned.
\end{abstract}

\section{Introduction}

Producing new methods by combining old ones is established stochastic
practice. Genetic programming assembles programs by mutation and
crossover~\cite{koza1992}, and evolutionary methods generate images and
designs the same way~\cite{sims1991}.

Language models now serve as the variation operator in the same loop:
FunSearch~\cite{funsearch2024}, AlphaEvolve~\cite{alphaevolve2025}, and
related work~\cite{lehman2022,meyerson2023}. All of these search a discrete
space of individual artifacts by sampling variations of them.

A separate tradition makes algorithm design continuous by tuning the
constants of a fixed template: algorithm selection~\cite{rice1976},
hyper-heuristics~\cite{burke2013}, automated configuration. There the space
is continuous but the architecture is frozen.

The inspiration simplex is a space that is continuous \emph{and}
architecturally open: barycentric coordinates over $K$ named methods, with
a language model mapping each coordinate to a working program. Performance
on a benchmark then defines an objective that standard derivative-free
machinery can search.

\section{The construction}

Fix $K$ established methods, the vertices. A candidate is a point
$w = (w_1, \dots, w_K)$ with $w_k \ge 0$ and $\sum_k w_k = 1$: a recipe
saying how much of each ancestor a new program should contain
(Figure~\ref{fig:simplex}).

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1.35]
  \coordinate (NM) at (0, 2.0);
  \coordinate (DE) at (-2.5, -0.45);
  \coordinate (PS) at (2.5, -0.45);
  \coordinate (SA) at (0, -2.0);
  \coordinate (CMA) at (0.55, 0.4);
  \draw[thick] (NM) -- (DE) -- (SA) -- (PS) -- cycle;
  \draw[thick] (NM) -- (SA);
  \draw[thick] (DE) -- (PS);
  \draw[thick, dashed, gray] (CMA) -- (NM);
  \draw[thick, dashed, gray] (CMA) -- (DE);
  \draw[thick, dashed, gray] (CMA) -- (PS);
  \draw[thick, dashed, gray] (CMA) -- (SA);
  \fill (NM) circle (1.1pt) node[above] {Nelder--Mead};
  \fill (DE) circle (1.1pt) node[left] {Differential Evolution};
  \fill (PS) circle (1.1pt) node[right] {Pattern Search};
  \fill (SA) circle (1.1pt) node[below] {Simulated Annealing};
  \fill[gray] (CMA) circle (1.0pt) node[above right, gray] {CMA-ES};
  \coordinate (W) at (0.11, -0.1);
  \fill (W) circle (1.3pt);
  \node[below left] at (W) {\scriptsize centroid $\rightarrow$ Alloy};
\end{tikzpicture}
\caption{The five-vertex simplex drawn in perspective: a $4$-simplex has
ten edges, and the dashed ones recede to the rear vertex (CMA-ES). The
marked interior point is the equal-weights centroid, the recipe whose
prompt (Figure~\ref{fig:prompt}) produced Alloy.}
\label{fig:simplex}
\end{figure}

The example uses five classical methods as vertices:
Nelder--Mead~\cite{neldermead1965}, Differential
Evolution~\cite{stornprice1997}, CMA-ES~\cite{hansen2001}, pattern
search~\cite{hookejeeves1961}, and simulated
annealing~\cite{kirkpatrick1983}. The generator is a frontier language
model (\texttt{claude-opus-4-8}).

Every iterative optimizer has the same five parts: how it starts
(initialization), how it proposes the next candidate (move generation), how
it decides to keep or reject it (acceptance), how it adjusts step sizes
(adaptation), and when it restarts (restart).

Each vertex algorithm is famous for some parts and silent on others.
Differential evolution is a way of initialising and proposing moves;
simulated annealing is a way of accepting moves and restarting.

The heart of the map from recipe to program is an asymmetric instruction.
The heaviest vertex is the host and supplies the architecture; every other
vertex is a graft, and ``70\% A, 30\% B'' means ``A, but borrow an idea
from B'', not a chimera. What B's idea looks like inside A's architecture
is the model's decision.

The weights also set quantitative anchors. Each part is divided among the
vertices that have a mechanism for it, in proportion to their weights, and
the shares appear in the prompt as percentages: a vertex holding 25\% of
move generation means its move must be chosen a quarter of the time.

The coordinate therefore acts on two scales. Coarse changes act
semantically: they change who hosts, which ideas appear, and how
prominently. Fine changes, 20\% versus 15\% of a graft, move only the
percentages; at that scale the map is parametric.

Figure~\ref{fig:prompt} shows the actual prompt at the equal-weights
center. The prompt is a deterministic function of the coordinate; the model
decides how to implement each mechanism and how to wire the parts together.

\begin{figure}[p]
\scriptsize
\begin{verbatim}
Build a black-box numerical optimizer by BLENDING base algorithms - but
blending is ASYMMETRIC, not a 50/50 average. The highest-weight method is
the HOST architecture; the others donate GRAFTED ideas in proportion to
their weight (a 30% method contributes a prominent borrowed mechanism; a 5%
method only a light inflection). '70% A, 30% B' means 'A, but borrow an
idea from B', not a chimera.

HOST architecture (20%): NelderMead - build the skeleton from this.
GRAFT into it: 20% DifferentialEvolution, 20% CMAEvolutionStrategy,
20% PatternSearch, 20% SimulatedAnnealing

Inspiration weights: NelderMead 20%, DifferentialEvolution 20%,
CMAEvolutionStrategy 20%, PatternSearch 20%, SimulatedAnnealing 20%

Per-slot guidance (host owns a slot unless a graft's weight dominates it):
  - initialization: DifferentialEvolution (100%): population +
    difference-vector mutation (rand/1, current-to-best/1) and crossover
  - move_generation: NelderMead (25%): downhill simplex: reflect/expand/
    contract/shrink a set of n+1 points; DifferentialEvolution (25%):
    population + difference-vector mutation and crossover;
    CMAEvolutionStrategy (25%): sample from a Gaussian whose covariance
    adapts to successful steps; PatternSearch (25%): Hooke-Jeeves
    coordinate moves with a step size that shrinks on failure
  - acceptance: SimulatedAnnealing (100%): accept uphill moves with
    probability exp(-dF/T); cool T over time
  - adaptation: NelderMead (33%); CMAEvolutionStrategy (33%);
    PatternSearch (33%)
  - restart: SimulatedAnnealing (100%)

Write connective glue as needed so the slots cooperate (e.g. the
acceptance rule gates the move-generation output; adaptation updates the
step/covariance used by move-generation). Favour a real, working blend
over a faithful copy of any single algorithm.

Write a single Python function with EXACTLY this signature and contract:

    def optimize(objective, n_trials, n_dim):
        """Minimise `objective` (a callable taking a list of n_dim floats
        in [0,1] and returning a float). Use AT MOST n_trials calls to
        objective. Return (best_value, best_point)."""

Hard rules:
  - Pure Python. You may `import math` and `import random` only.
  - Every candidate must be clipped into [0,1] before evaluation.
  - Count objective calls; never exceed n_trials.
  - No global state, no prints. Just the function.

Return ONLY a ```python code block containing the optimize function.
\end{verbatim}
\caption{The prompt at the equal-weights point
$w = (0.2, 0.2, 0.2, 0.2, 0.2)$, lightly line-wrapped for the page. Moving
$w$ changes the host, the graft proportions, and the per-slot shares, and
nothing else.}
\label{fig:prompt}
\end{figure}

The generated program is compiled, validated, and scored on a benchmark,
and that score is the objective value at $w$. Generation is stochastic, so
the objective is noisy, as simulation-based objectives usually are.

Because the simplex maps to the unit cube by a standard bijection, any
off-the-shelf derivative-free optimizer can search it; the outer loop sees
only a box-constrained noisy objective.

The geometry protects the search. Every point is a committee of mechanisms
that individually work, so the worst case of a step is a mediocre blend of
good algorithms, not an arbitrary program; and nearby points receive nearby
specifications, which gives a derivative-free optimizer something to
exploit.

\section{First example: Alloy}

Candidates are scored on disguised versions of real-world objectives
(enzyme kinetics, wind-farm layout, battery dispatch, and others). A seeded
smooth bijection of the unit cube relocates each optimum, so a program
cannot succeed by memorisation.

The score is regret normalised against a panel of standard optimizers,
averaged over problems and seeds: 0 matches the best panel member
everywhere, 1 the worst. Selection used sixteen problems, three seeds, 120
evaluations per run.

\subsection{Selection}

Table~\ref{tab:warm} scores the five pure recipes and the equal-weights
centroid. The centroid's program, named Alloy, wins at 0.087; the best pure
vertex, Nelder--Mead, scores 0.160.

\begin{table}[h]
\centering
\begin{tabular}{lc}
\toprule
recipe & regret \\
\midrule
centroid (equal fifths) $\rightarrow$ Alloy & 0.087 \\
pure Nelder--Mead & 0.160 \\
pure Simulated Annealing & 0.225 \\
pure Differential Evolution & 0.358 \\
pure Pattern Search & 0.477 \\
pure CMA-ES & 1.000 \\
\bottomrule
\end{tabular}
\caption{Panel-normalised regret (lower is better) of the program generated
at each recipe: sixteen disguised objectives, three seeds, 120 evaluations
per run.}
\label{tab:warm}
\end{table}

Structure mattered at generation time. Eight free-form attempts by the same
model, prompted simply for the strongest optimizer it could write, gave a
best of 0.141 and a median of 0.202 on the identical suite, against Alloy's
0.087.

\subsection{Validation on problems the search never saw}
\label{sec:heldout}

A selection winner can be a fluke, so the test that matters is out of
sample. Twenty-nine objectives, spanning 2 to 90 dimensions, were verified
against the full history of the benchmark library to have been used by no
selection step; each ran at budgets of 60, 120, 240 and 480 evaluations
with five seeds, 580 instances in all.

Alloy has the best mean rank and the most outright wins at every budget
against six competitors: Nelder--Mead, Differential Evolution, CMA-ES,
Nevergrad's CMA, a tuned 14-gene template, and the best free-form
generation. Pairwise it beats each of them on 64\% to 77\% of instances,
with sign-test $p$-values from $4\times10^{-11}$ down to
$1\times10^{-40}$.

Nor is it a lucky singleton. Adding the other strong selection-stage blends
gives Table~\ref{tab:untouched}: the blends hold the top four places at the
two smaller budgets, above every classical implementation, and the best of
them is statistically even with Alloy (48\% pairwise, $p = 0.29$).

\begin{table}[h]
\centering
\begin{tabular}{lcccc}
\toprule
 & \multicolumn{4}{c}{mean rank of 11 (lower is better)} \\
optimizer & 60 & 120 & 240 & 480 \\
\midrule
Alloy & 4.6 & 4.3 & 4.1 & 4.4 \\
warm4 blend (Patt 53, CMA 23) & 4.0 & 4.2 & 4.8 & 5.7 \\
rand14 blend (Patt 64, CMA 21) & 4.0 & 4.6 & 5.6 & 6.4 \\
14-gene template (evolved) & 5.8 & 5.6 & 5.3 & 5.4 \\
warm7 blend (Patt 92) & 5.0 & 5.1 & 5.9 & 7.4 \\
generated pure Nelder--Mead & 5.5 & 6.2 & 6.7 & 6.1 \\
Nevergrad CMA & 6.7 & 6.6 & 5.9 & 5.7 \\
Differential Evolution & 7.6 & 7.0 & 6.4 & 5.2 \\
CMA-ES & 7.7 & 7.1 & 6.3 & 5.7 \\
free-form best & 6.8 & 7.3 & 6.9 & 6.4 \\
Nelder--Mead & 7.8 & 7.8 & 7.7 & 7.1 \\
\bottomrule
\end{tabular}
\caption{Race on the 29 held-out problems, 580 instances per budget column,
rows ordered by mean rank across budgets. Blends dominate at small budgets;
Alloy is the most consistent.}
\label{tab:untouched}
\end{table}

Alloy initialises with a DE population and builds a Nelder--Mead simplex
from the best members. At each step it picks one of four move generators
with equal probability (simplex reflect-expand-contract, DE mutation with
crossover, a Gaussian perturbation with success-adapted covariance, a
coordinate probe with shrinking step); every candidate passes a
simulated-annealing acceptance gate, and stagnation triggers a reheat
around the incumbent. It ships in the humpday package, in Python and
JavaScript.

A distilled twin performs identically. Rewriting Alloy's architecture as
one ordinary program with the mixing proportions as parameters, then tuning
those with twenty cheap evaluations, gives an optimizer statistically even
with Alloy on the held-out race (45\% pairwise, $p = 0.05$) with no
language model anywhere at run time.

\section{What did not work}
\label{sec:negative}

Generation is a lottery. Five draws at the centroid recipe scored 0.087,
0.249, 0.301, 0.320 and 1.000, and roughly one generation in seven fails to
compile. Alloy is the validated draw, not the expected one; the procedure
that works samples a few draws and keeps what validates.

The map has blind spots. The pure CMA-ES recipe never produced a working
program: correct covariance adaptation is too much to write in budgeted
pure Python, and its generations were degenerate.

The continuous search never earned its keep on this vertex set. The winner
was the centroid, the one point anyone would evaluate first; random sweeps
and a guided derivative-free outer loop (best-of-twenty 0.270 against
0.308 for random sampling) found nothing better. A second round with
unequal vertices confirmed the diagnosis: there the equal blend was
mid-pack, so the centroid's win reflects a symmetric vertex set, not a
generally rich interior.

In-sample scores mislead, reliably. A 14-gene parametric template tuned by
evolutionary search posted the best selection score of anything run (0.244)
and then collapsed to mean rank 7 of 13 on the held-out problems. Free-form
one-shots sit at the bottom of every held-out table despite respectable
selection scores. Only races on untouched problems told the truth.

Cheap parametric search matches expensive semantic search per evaluation.
On identical coordinates, a hand-parameterized version of Alloy's
architecture scored 0.272 in-sample where fresh language-model generations
scored 0.308. The model's value was the architecture, found once; after
distillation it is no longer needed.

\section{Seen from global optimization}

The two-level structure is classical, split across two spaces. The semantic
move is the basin hop and the parametric tuning is the local
descent~\cite{wales1997}; done inside the loop it is memetic
search~\cite{moscato1989}; and the simplex is a continuous relaxation of
architecture choice, as in differentiable architecture
search~\cite{liu2019darts}, with a language model realizing the mixture
instead of a weighted sum.

The closest workflow is SHGO~\cite{endres2018}: sample the domain, identify
the distinct basins, run one local minimization per basin. The basins live
in the space of generated programs, not in $w$, since one coordinate yields
different architectures across draws.

The recipe that survives: generate a small number of structured blends,
validate the winner on problems the search never touched, distill it to a
parametric family, and tune cheaply from then on. The language model is
used a handful of times, early, and never again.

\section{Outside the simplex: negative weights}
\label{sec:outside}

Drop the non-negativity constraint and keep the sum: recipes now range over
an affine slice of which the simplex is the all-positive face. A negative
weight needs a semantics, and six were tested. Avoidance: shadow the
negated algorithm's proposals at zero cost and steer away. Density: thin
the host's proposals toward the signed mixture of proposal densities.
Caricature: exaggerate the host along the axes where it most differs from
the negated reference. Pitfalls: design the negated algorithm's failure
modes out of the host. Baseline: treat its shadow as the bar to beat.
Residual: graft in, amplified, only what the negated algorithm has that the
host lacks.

Attribution is by paired ablation. Every generated program gates its
negative-weight machinery through one module constant, so its twin, the
same program with the constant zeroed, recovers the plain host and is
scored on identical instances. Single-draw leaderboards reward whatever a
label elicits from the generator; only the paired delta counts.

The protocol exists because avoidance sets the trap. Its best recipe won a
selection round outright (0.179 against 0.326 for the best unsigned
control), and disabling the repulsion in that same program improved it to
0.130: the win belonged to the unusually good host the prompt elicited, and
the signed term was a handicap.

Under the protocol, caricature is the only semantics of the six that helps
on average at selection (Table~\ref{tab:sem}), and the mathematically
prettiest, rejection-thinned proposal densities, goes zero for five against
its twins. Sweeping all twenty ordered pairs, eight of nineteen benefit
from caricature, the map is uncorrelated with the ancestors' rank
correlations ($+0.05$), and the magnitude dial optimises at restraint: mild
exaggeration helps, twofold and beyond tips into parody.

\begin{table}[h]
\centering
\begin{tabular}{lcc}
\toprule
semantics & mean paired delta & arms improving \\
\midrule
caricature & $-0.032$ & 4/8 \\
repulsion  & $-0.009$ & 3/7 \\
residual   & $+0.014$ & 2/8 \\
pitfalls   & $+0.018$ & 2/7 \\
density    & $+0.040$ & 0/5 \\
baseline   & $+0.086$ & 0/4 \\
\bottomrule
\end{tabular}
\caption{Six semantics for a negative weight under the paired protocol,
selection suite spanning 2 to 100 dimensions; negative delta means the
signed machinery helps.}
\label{tab:sem}
\end{table}

Searching the extended slice (weights in roughly $[-1,2]$, each negative
weight a caricature reference with its magnitude as exaggeration length),
the guided outer loop reached 0.078 with two thirds of its evaluations
improving on their twins, against 0.047, 0.198 and 0.374 for three random
shards: one lucky draw aside, guidance pays. That lucky 0.047, paired delta
$-0.134$ at selection, sets the second trap: on the held-out problems its
twin beats it on 65\% of paired instances ($p = 9\times10^{-12}$). The
paired ablation solves attribution, not transfer; a delta earned on the
selection distribution is a claim about that distribution only, so the
protocol acquires a second stage, and only deltas that survive out of
sample count.

Two do. The CMA-ES-away-from-DE artifact beats its twin on 65\% of held-out
paired instances ($p = 2\times10^{-12}$), and annealing-away-from-pattern-%
search on 59\% ($p = 10^{-5}$). The pattern across the raced winners is
consistent: where the twin is already excellent the signed term vanishes or
reverses; where the twin is weak the signed term repairs it. Negative
weights, under the one semantics that works, are a treatment for weak
hosts, not an enhancement for strong ones, and no signed artifact yet leads
the held-out field absolutely.

A byproduct, twice replicated: the twins of the strongest signed recipes,
budget-aware pattern-search hosts with Nelder--Mead grafts, rank first of
fourteen at the largest budget on the held-out problems, ahead of every
prior artifact. Exploring outside the simplex keeps producing better
inhabitants of the inside.

\section{Generalization and further examples}

Nothing here is specific to optimizers. The construction needs named
concepts the model knows, a decomposition of the target artifact into
parts, and a measurable score. Sorting algorithms, data structures, trading
rules, prompt styles, and agent architectures all qualify, and further
worked examples are planned for later versions of this paper.

The construction also refines. Like SHGO one can subdivide, searching the
sub-simplex spanned by the best recipes at finer resolution; unlike SHGO
one can promote discovered programs to vertices, so each generation's
validated winners become the next generation's ancestors while the geometry
stays $K$-dimensional.

The boundaries of the claims: budgets are small, the regime where blending
plausibly helps most, and one generator model was used. The suite, while
disguised, derives from public examples, so relatives of these objectives
are likely in the generator's training data even though the optima's
locations are not.

\vspace{1em}
\noindent\emph{Availability.} Alloy ships in the humpday
package~\cite{humpday}, in Python and JavaScript. The search harnesses, the
run logs behind every number in this paper, and every generated program are
in the same repository.

\begin{thebibliography}{9}

\bibitem{humpday} P.~Cotton.
HumpDay: recommendations and comparisons of derivative-free optimizers.
\texttt{https://humpday.microprediction.org}, 2026.

\bibitem{koza1992} J.~R. Koza.
\emph{Genetic Programming: On the Programming of Computers by Means of
Natural Selection}. MIT Press, 1992.

\bibitem{sims1991} K.~Sims.
Artificial evolution for computer graphics.
\emph{Computer Graphics (SIGGRAPH)}, 25(4):319--328, 1991.

\bibitem{funsearch2024} B.~Romera-Paredes et al.
Mathematical discoveries from program search with large language models.
\emph{Nature}, 625:468--475, 2024.

\bibitem{alphaevolve2025} A.~Novikov et al.
AlphaEvolve: a coding agent for scientific and algorithmic discovery.
Google DeepMind technical report, 2025.

\bibitem{lehman2022} J.~Lehman, J.~Gordon, S.~Jain, K.~Ndousse, C.~Yeh, and
K.~O. Stanley. Evolution through large models.
arXiv:2206.08896, 2022.

\bibitem{meyerson2023} E.~Meyerson, M.~J. Nelson, H.~Bradley, A.~Gaier,
A.~Moradi, A.~K. Hoover, and J.~Lehman.
Language model crossover: variation through few-shot prompting.
arXiv:2302.12170, 2023.

\bibitem{rice1976} J.~R. Rice.
The algorithm selection problem.
\emph{Advances in Computers}, 15:65--118, 1976.

\bibitem{burke2013} E.~K. Burke, M.~Gendreau, M.~Hyde, G.~Kendall, G.~Ochoa,
E.~\"Ozcan, and R.~Qu.
Hyper-heuristics: a survey of the state of the art.
\emph{Journal of the Operational Research Society}, 64(12):1695--1724, 2013.

\bibitem{neldermead1965} J.~A. Nelder and R.~Mead.
A simplex method for function minimization.
\emph{The Computer Journal}, 7(4):308--313, 1965.

\bibitem{stornprice1997} R.~Storn and K.~Price.
Differential evolution: a simple and efficient heuristic for global
optimization over continuous spaces.
\emph{Journal of Global Optimization}, 11:341--359, 1997.

\bibitem{hansen2001} N.~Hansen and A.~Ostermeier.
Completely derandomized self-adaptation in evolution strategies.
\emph{Evolutionary Computation}, 9(2):159--195, 2001.

\bibitem{hookejeeves1961} R.~Hooke and T.~A. Jeeves.
Direct search solution of numerical and statistical problems.
\emph{Journal of the ACM}, 8(2):212--229, 1961.

\bibitem{kirkpatrick1983} S.~Kirkpatrick, C.~D. Gelatt, and M.~P. Vecchi.
Optimization by simulated annealing.
\emph{Science}, 220(4598):671--680, 1983.

\bibitem{wales1997} D.~J. Wales and J.~P.~K. Doye.
Global optimization by basin-hopping and the lowest energy structures of
Lennard-Jones clusters.
\emph{Journal of Physical Chemistry A}, 101(28):5111--5116, 1997.

\bibitem{moscato1989} P.~Moscato.
On evolution, search, optimization, genetic algorithms and martial arts:
towards memetic algorithms.
Caltech C3P Report 826, 1989.

\bibitem{liu2019darts} H.~Liu, K.~Simonyan, and Y.~Yang.
DARTS: differentiable architecture search.
\emph{International Conference on Learning Representations}, 2019.

\bibitem{endres2018} S.~C. Endres, C.~Sandrock, and W.~W. Focke.
A simplicial homology algorithm for Lipschitz optimisation.
\emph{Journal of Global Optimization}, 72:181--217, 2018.

\end{thebibliography}

\end{document}
